---
title: "Development, internal-external validation, and use of a prognostic model to predict future foot complications among people with diabetes recently discharged from hospital in Ontario, Canada"
always_allow_html: true
output:
  github_document:
    toc: true
    toc_depth: 5
  keep_text: true
  pandoc_args: --webtex
---

## Overview
  
This file contains R code illustrating how to:

1. Obtain all relevant model information (coefficients, hyperparameters etc.) from the model objects saved in this repository.
2. Use the model objects to generate predictions on new test data.
3. Assess model performance on the new test data.

## Set-up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE
)
```

To run the code below, please install and load the following R libraries:

```{r}
library(riskRegression)
library(randomForestSRC)
library(rms)
library(prodlim)
```


## Models


The final models (trained on the whole cohort reported in Roberst & Loeffler et al.) are saved as .rds files in this repository and can be loaded as follows: 

```{r}
## load model objects
FGR <- readRDS("models/final_FGR_clean.rds") # Fine-Gray Regression
# RSFCR <-

```

### Fine-Gray Regression (FGR)

The FGR model object contains the following items:

```{r}
names(FGR)
# crrFit = model fit obtained with riskRegression::FGR()
# call = original model call (note: variance was set to FALSE, SEs of coefficients were obtained using bootstrapping)
# terms = model terms
# form = model formula
# cause = 1 (event of interest = 1 = foot complication)
# coef = optimized model coefficients (SEs) and corresponding subdistribution hazard ratios (sHR)
# splines = knot locations for restricted cubic splines for continuous predictors
```

### Random Survival Forest for Competing Risks (RSFCR)

The RSFCR model object


## Dummy data

```{r}

```

## Generate new predictions

```{r}
## create new test data
new_data <- data.frame(
  age = 67,
  sex_f = 1,
  elective_adm = 1,
  homelessness = 0,
  peripheral_AD = 0,
  coronary_AD = 1,
  stroke = 0,
  CHF = 0,
  hypertension = 1,
  COPD = 0,
  CKD = 0,
  malignancy = 0,
  mental_illness = 0,
  creatinine = 140.0,
  Hb_A1C = 8.5,
  albumin = 32.1,
  Hb_A1C_missing = 0,
  creatinine_missing = 0,
  albumin_missing = 0
)

## add splines to data
# derive splines based on knot locations
age_splines <- rcs(new_data$age, FGR$splines$age_knots)
creatinine_splines <- rcs(new_data$creatinine, FGR$splines$creatinine_knots)
Hb_A1C_splines <- rcs(new_data$Hb_A1C, FGR$splines$hba1c_knots)
albumin_splines <- rcs(new_data$albumin, FGR$splines$albumin_knots)

## add non-linear components to new_data
new_data$age1 <- age_splines[, 2]
new_data$age2 <- age_splines[, 3]
new_data$creatinine1 <- creatinine_splines[, 2]
new_data$creatinine2 <- creatinine_splines[, 3]
new_data$Hb_A1C1 <- Hb_A1C_splines[, 2]
new_data$Hb_A1C2 <- Hb_A1C_splines[, 3]
new_data$albumin1 <- albumin_splines[, 2]

# predicted risk at 1 year
predict(FGR, newdata = new_data, times = 365.25)[1] # 0.0171758

# # to plot CIF, we need to extract predicted values at multiple time points
# time <- seq(1, 365 * 5, 5) # predict up to 5 years in steps of 5 days
# p <- predict(model, newdata = new_data, times = time)
# plot(time, p, type = "l")
```